---
title: "Advanced Features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {.alert .alert-info}
🇫🇷 **[Version française disponible](advanced-features_FR.html)** / **French version available**

📖 **See also:** [Getting Started Guide](getting-started.html) for basic usage
:::

This vignette covers advanced features of `quartify` including customization options, batch processing, Quarto Books, code quality integration, and special features.

## Customization Options

You can customize the output document with several parameters:

```{r eval=FALSE}
library(quartify)

rtoqmd(
  input_file = "my_script.R",
  output_file = "my_document.qmd",
  title = "My Analysis Report",
  author = "Your Name",
  format = "html",
  theme = "cosmo",
  render_html = TRUE,
  open_html = TRUE,
  code_fold = TRUE,
  number_sections = TRUE,
  show_source_lines = TRUE
)
```

### Available Parameters

- **`input_file`**: Path to your R script
- **`output_file`**: Path for the output Quarto document (optional)
- **`title`**: Title for the document (default: "My title")
- **`author`**: Author name (default: "Your name")
- **`format`**: Output format - always "html" (parameter kept for backward compatibility)
- **`theme`**: Quarto HTML theme (default: "cosmo")
- **`render_html`**: Whether to render the .qmd to HTML (default: TRUE)
- **`output_html_file`**: Custom path for HTML output (optional)
- **`open_html`**: Whether to open the HTML file after rendering (default: FALSE)
- **`code_fold`**: Whether to fold code blocks by default (default: FALSE)
- **`number_sections`**: Whether to number sections automatically (default: TRUE)
- **`show_source_lines`**: Show original line numbers in code chunks (default: TRUE)
- **`lang`**: Language for UI elements ("en" or "fr", default: "en")

### HTML Themes

`quartify` supports 25+ Quarto themes for HTML output:

**Light themes**: cosmo, flatly, litera, lumen, lux, materia, minty, morph, pulse, quartz, sandstone, simplex, sketchy, spacelab, united, yeti, zephyr

**Dark themes**: cyborg, darkly, slate, solar, superhero, vapor

**Combined**: cerulean (light-dark)

Try different themes to find the perfect look:

```{r eval=FALSE}
# Modern and clean
rtoqmd("script.R", theme = "cosmo")

# Dark mode
rtoqmd("script.R", theme = "darkly")

# Academic style
rtoqmd("script.R", theme = "litera")
```

## Batch Converting Directories

### Using rtoqmd_dir()

Convert all R scripts in a directory (including subdirectories) programmatically:

```{r eval=FALSE}
# Convert all R scripts in a directory
rtoqmd_dir("path/to/scripts")

# Convert and render to HTML
rtoqmd_dir("path/to/scripts", render_html = TRUE)

# With custom author and title prefix
rtoqmd_dir("path/to/scripts", 
           title_prefix = "Analysis: ",
           author = "Data Team")

# Exclude certain files (e.g., test files)
rtoqmd_dir("path/to/scripts", 
           exclude_pattern = "test_.*\\.R$")

# Non-recursive (current directory only)
rtoqmd_dir("path/to/scripts", recursive = FALSE)
```

### Directory Mode in Shiny Apps

The most intuitive way to batch convert files is through the **Directory Mode** in either the RStudio addin or standalone app:

```{r eval=FALSE}
# Launch the standalone app
quartify_app()

# Or use the RStudio addin: Addins > "Convert R to Quarto (rtoqmd)"
```

**Directory Mode features:**

1. **Browse for directory**: Select the folder containing your R scripts
2. **Recursive option**: Include or exclude subdirectories
3. **Output directory**: Optionally specify a different location for converted files
4. **Create Book checkbox**: Automatically generate `_quarto.yml` to combine all documents into a Quarto book
5. **Render option**: Choose whether to generate HTML files for all scripts
6. **One-click conversion**: Click the GENERATE button to convert all files

## Creating Quarto Books

One of the most powerful features of `quartify` is the ability to automatically create Quarto books from directories of R scripts.

### Automatic Book Creation

When using `rtoqmd_dir()` with the `create_book` parameter, quartify automatically generates a `_quarto.yml` configuration file:

```{r eval=FALSE}
# Convert directory and create a Quarto book
rtoqmd_dir(
  dir_path = "my_scripts/",
  output_html_dir = "documentation/",
  create_book = TRUE,
  book_title = "My Project Documentation",
  render_html = TRUE
)
```

This creates:

```
documentation/
├── _quarto.yml           # Book configuration
├── index.qmd             # Book landing page
├── script1.qmd
├── script2.qmd
├── subfolder/
│   └── script3.qmd
└── _book/                # Rendered HTML book (if render_html = TRUE)
    ├── index.html
    ├── script1.html
    ├── script2.html
    └── ...
```

### Book Features

The generated Quarto book includes:

- **Automatic navigation**: Sidebar with all chapters organized by directory structure
- **Search functionality**: Built-in search across all documents
- **Cross-references**: Link between documents within the book
- **Unified styling**: Consistent theme across all documents
- **Table of contents**: Per-chapter and book-level navigation

### Manual Book Configuration

You can also manually configure a Quarto book. Create a `_quarto.yml` file:

```yaml
project:
  type: book

book:
  title: "My Analysis Collection"
  author: "Your Name"
  chapters:
    - index.qmd
    - chapter1.qmd
    - chapter2.qmd

format:
  html:
    theme: cosmo
    code-fold: false
```

Then convert your scripts to the book directory and render:

```{r eval=FALSE}
# Convert scripts to book directory
rtoqmd_dir("scripts/", output_dir = "book/", render_html = FALSE)

# Render the book
quarto::quarto_render("book/")
```

## Code Quality Integration

`quartify` integrates with **styler** and **lintr** to provide automated code quality checks and formatting suggestions directly in your generated documentation. This is particularly useful for:

- **Teaching materials**: Show students both original and properly formatted code
- **Code review**: Automatically identify style issues in documentation
- **Best practices**: Learn tidyverse style guide by example
- **Quality control**: Ensure consistent code style across projects

### Features Overview

- **`use_styler = TRUE`**: Shows formatted code according to the [tidyverse style guide](https://style.tidyverse.org/)
- **`use_lintr = TRUE`**: Identifies code quality issues and potential problems
- **`apply_styler = TRUE`**: Directly applies styling to your original R script (⚠️ modifies source file)

When quality issues are detected, `quartify` creates **interactive tabsets** in the HTML output with:

- **Original Code**: Your code as written
- **Styled Code**: Formatted version (only shown if changes are needed)
- **Lint Issues**: Quality warnings (only shown if issues are found)

**Smart Tabsets**: Tabsets are only created when needed. If there are no style changes or lint issues, you get a regular code chunk (no tabset).

### Installation

Code quality features require optional dependencies:

```{r eval=FALSE}
install.packages(c("styler", "lintr"))
```

These packages are only loaded when you use the code quality features.

### Basic Usage

#### Show Styling Suggestions

```{r eval=FALSE}
# Show styled code in tabsets
rtoqmd("script.R", "output.qmd", 
       use_styler = TRUE,
       render_html = TRUE)
```

This generates tabsets showing both the original and styled code side-by-side.

#### Show Lint Issues

```{r eval=FALSE}
# Show code quality issues
rtoqmd("script.R", "output.qmd", 
       use_lintr = TRUE,
       render_html = TRUE)
```

This adds a "Lint Issues" tab with detailed quality feedback.

#### Combined Quality Checks

```{r eval=FALSE}
# Show both styling and lint issues
rtoqmd("script.R", "output.qmd",
       use_styler = TRUE,
       use_lintr = TRUE,
       render_html = TRUE,
       open_html = TRUE)
```

This creates comprehensive quality reports with three tabs:
1. Original Code
2. Styled Code
3. Lint Issues

### Applying Styles to Source Files

::: {.alert .alert-warning}
**⚠️ Warning**: `apply_styler = TRUE` **modifies your original R script file**. Always use version control or create backups before using this option.
:::

```{r eval=FALSE}
# Apply styling directly to source file
rtoqmd("my_script.R", "output.qmd",
       use_styler = TRUE,
       apply_styler = TRUE)
```

**Important**: When `apply_styler = TRUE`:

1. Your source R file is modified before conversion
2. No tabsets are created (source is already styled)
3. `use_styler` must also be TRUE
4. The conversion uses the styled source file

This is useful for:

- Cleaning up code before committing to version control
- Batch formatting multiple scripts
- Preparing code for publication or teaching

### Example with Code Quality

Try the included example that demonstrates code quality features:

```{r eval=FALSE}
# Locate the example file
example_file <- system.file("examples", "example_code_quality.R", 
                           package = "quartify")

# View the file content to see intentional style issues
file.show(example_file)

# Convert with both checks
rtoqmd(example_file, "quality_demo.qmd", 
       use_styler = TRUE, 
       use_lintr = TRUE,
       render_html = TRUE,
       open_html = TRUE)
```

The example file includes intentional style issues:

```r
# Here we have some code with style issues
x = 3  # Should use <- instead of =
y <- 2
z<-10  # Missing spaces around <-

# Calculate statistics
mean_value<-mean(c(x,y,z)) # Missing spaces

# Create a vector with inconsistent spacing
my_vector<-c(1,2,3,4,5)
```

### Example Output

Given code with style issues:

```r
x = 3  # Should use <- instead of =
y <- 2
z<-10  # Missing spaces
```

With `use_styler = TRUE` and `use_lintr = TRUE`, you'll see a tabset:

**Original Code tab:**
```r
x = 3  # Should use <- instead of =
y <- 2
z<-10  # Missing spaces
```

**Styled Code tab:**
```r
x <- 3 # Should use <- instead of =
y <- 2
z <- 10 # Missing spaces
```

**Lint Issues tab:**
- Line 1: Use `<-` or `<<-` for assignment, not `=`.
- Line 3: Put spaces around all infix operators.

### Batch Processing with Quality Checks

Apply code quality checks to entire directories:

```{r eval=FALSE}
# Convert directory with quality checks
rtoqmd_dir(
  dir_path = "my_scripts/",
  output_html_dir = "documentation/",
  use_styler = TRUE,
  use_lintr = TRUE,
  render_html = TRUE
)
```

This applies code quality checks to every R script in the directory.

### Using in Shiny Apps

All three Shiny applications (`rtoqmd_addin()`, `quartify_app()`, and `quartify_app_web()`) include checkboxes for code quality options:

- ☑️ **Use styler formatting** (shows styled version in tabs)
- ☑️ **Use lintr quality checks** (shows issues in tabs)
- ☑️ **Apply styler to source file** (modifies original R file) - *Not available in web version*

Simply check the desired options before clicking GENERATE.

### Benefits of Code Quality Integration

1. **Learn by Example**: See proper R coding style alongside your code
2. **Code Review**: Automatic quality feedback without manual review
3. **Teaching Tool**: Perfect for educational materials and tutorials
4. **Documentation**: Generate clean, well-styled code in documentation
5. **Best Practices**: Enforce tidyverse style guide automatically
6. **Consistency**: Maintain uniform code style across projects

### Configuration

Both styler and lintr use their default configurations. For custom settings:

- **styler**: Create a `.styler.R` file in your project root
- **lintr**: Create a `.lintr` file in your project root

See their respective documentation:
- [styler documentation](https://styler.r-lib.org/)
- [lintr documentation](https://lintr.r-lib.org/)

### Performance Notes

- Code quality checks add minimal processing time (typically < 1 second per chunk)
- Checks only run when parameters are TRUE (default: FALSE)
- Suitable for interactive and automated workflows
- For very large projects, consider selective conversion

### Troubleshooting

**Packages not found:**

```{r eval=FALSE}
install.packages("styler")  # For formatting
install.packages("lintr")   # For linting
```

**Checks fail for specific chunks:**

If code quality checks fail for a particular code chunk, the original code is displayed without a tabset, and a warning is logged. The conversion continues normally.

## Special Features

### Mermaid Diagrams

`quartify` supports [Mermaid](https://mermaid-js.github.io/) diagrams in your R scripts. Use special comments to include diagrams in your documentation:

```r
# mermaid-start
# graph TD
#   A[Start] --> B{Is it working?}
#   B -->|Yes| C[Great!]
#   B -->|No| D[Debug]
#   D --> A
# mermaid-end
```

This renders as a flowchart in your Quarto document.

### Callouts

Create callouts in your R scripts using special comment syntax:

```r
# callout-note - Important Note
# This is a note callout
# It can span multiple lines

# callout-warning - Be Careful
# This is a warning callout

# callout-tip - Pro Tip
# This is a tip callout
```

Supported callout types: `note`, `warning`, `tip`, `important`, `caution`

### RStudio Snippets

Install useful code snippets for faster R script writing:

```{r eval=FALSE}
install_quartify_snippets()
```

After installation and restarting RStudio, you can use:

- **`header`** + Tab: Insert document metadata template
- **`callout`** + Tab: Insert callout template
- **`mermaid`** + Tab: Insert Mermaid diagram template
- **`tabset`** + Tab: Insert tabset template

## Tips and Best Practices

### Structuring Your Scripts

1. **Start with structure**: Define your section headers first to create the document outline
2. **Use consistent levels**: Follow a logical hierarchy (2 → 3 → 4, don't skip levels)
3. **Add explanatory text**: Use regular comments to explain what your code does
4. **Group related code**: Keep related operations together
5. **Navigate easily**: In RStudio, use the document outline (Ctrl+Shift+O) to see your structure

### Code Quality

6. **Use code quality checks for teaching**: Enable `use_styler` and `use_lintr` for educational materials
7. **Learn from styled code**: Review the "Styled Code" tabs to learn best practices
8. **Fix issues incrementally**: Use lint feedback to improve code quality over time
9. **Version control before apply_styler**: Always commit before using `apply_styler = TRUE`

### Batch Conversion

10. **Use Directory Mode for projects**: Convert entire project folders with the Create Book option
11. **Exclude test files**: Use `exclude_pattern` to skip test scripts in batch conversion
12. **Organize by subdirectories**: Leverage recursive conversion to maintain project structure
13. **Set output directories**: Keep source R scripts separate from generated documentation

### Documentation Workflow

14. **Comment liberally**: More comments = better documentation
15. **Use metadata**: Add Title, Author, Date, and Description at the top of scripts
16. **Test incrementally**: Start with a small script to understand conversion behavior
17. **Preview frequently**: Use `render_html = TRUE` and `open_html = TRUE` to see results immediately

### CI/CD Integration

18. **Automate documentation**: Use GitHub Actions or GitLab CI to generate docs on commit
19. **Version control .R only**: Generate .qmd and .html files in CI/CD, not in your repository
20. **Deploy to GitHub Pages**: Automatically publish documentation with each push

## Next Steps

- **Basic Usage**: See [Getting Started Guide](getting-started.html) for fundamentals
- **French Version**: [Version française](advanced-features_FR.html)
- **GitHub Repository**: [ddotta/quartify](https://github.com/ddotta/quartify)
- **Report Issues**: [GitHub Issues](https://github.com/ddotta/quartify/issues)

## Conclusion

`quartify`'s advanced features enable sophisticated documentation workflows from code quality checks to automated book generation. Experiment with these features to find the best workflow for your projects!
