---
title: "Getting Started with quartify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with quartify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {.alert .alert-info}
🇫🇷 **[Version française disponible](getting-started_FR.html)** / **French version available**
:::

## Introduction

`quartify` is an R package that automatically converts R scripts into Quarto markdown documents (.qmd). The package recognizes RStudio code sections to create properly structured documents with navigation. This vignette will guide you through the basic usage and features of the package.

## Installation

You can install the development version of quartify from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ddotta/quartify")
```

## Basic Usage

The main function of the package is `rtoqmd()`. Here's a simple example:

```{r eval=FALSE}
library(quartify)

# Convert an R script to a Quarto document and render to HTML
rtoqmd("my_script.R", "my_document.qmd")

# Convert only, without rendering to HTML
rtoqmd("my_script.R", "my_document.qmd", render = FALSE)
```

By default, `rtoqmd()` will:
1. Create the .qmd file
2. Render it to HTML using Quarto
3. Open the HTML file in your default browser

### Output File Locations

Understanding where files are created:

- **If you specify an `output_file` path**: The .qmd file is created at that exact location, and the .html file (if rendered) is created in the same directory with the same base name.

```{r eval=FALSE}
# Creates: /path/to/my_analysis.qmd and /path/to/my_analysis.html
rtoqmd("script.R", "/path/to/my_analysis.qmd")
```

- **If you don't specify `output_file`**: The .qmd file is created in the same directory as your input R script, with the `.R` extension replaced by `.qmd`.

```{r eval=FALSE}
# If script.R is in /home/user/scripts/
# Creates: /home/user/scripts/script.qmd and /home/user/scripts/script.html
rtoqmd("/home/user/scripts/script.R")
```

- **Relative paths**: If you use relative paths, files are created relative to your current working directory (check with `getwd()`).

```{r eval=FALSE}
# Creates files in your current working directory
rtoqmd("script.R", "output.qmd")
```

## Structuring Your R Script

For optimal conversion, you need to follow specific commenting rules in your R script. `quartify` recognizes three types of lines:

### 0. Document Metadata (Optional)

You can define your document's metadata directly in your R script using special comments at the beginning of the file. These metadata will appear in the YAML header of the generated Quarto document.

**Recognized metadata:**

- **Title**: `# Title : My title` or `# Titre : Mon titre`
- **Author**: `# Author : My name` or `# Auteur : Mon nom`
- **Date**: `# Date : 2025-11-28`
- **Description**: `# Description : Description of your script`

::: {.alert .alert-success}
**Tip - RStudio Snippets:** To save time, you can create an [RStudio snippet](https://docs.posit.co/ide/user/ide/guide/productivity/snippets.html) to automatically insert this metadata header. Add this snippet to your R snippets (Tools > Edit Code Snippets > R):

```
snippet header
	# Title : ${1}
	#
	# Author : ${2}
	#
	# Date : ${3}
	#
	# Description : ${4}
	#
```

Once defined, type `header` followed by `Tab` in your R script to automatically insert the metadata structure.
:::

**Complete example with metadata:**

```r
# Title : Iris Data Analysis
#
# Author : Jane Doe
#
# Date : 2025-11-28
#
# Description : Explore differences between iris species
#

library(dplyr)

## Descriptive Analysis ####

# Calculate statistics by species

iris |> 
  group_by(Species) |>
  summarize(mean_sepal = mean(Sepal.Length))
```

**Behavior:**

- If metadata is found in the script, it **overrides** the `rtoqmd()` function parameters
- If no metadata is found, the function's `title` and `author` parameters are used
- Metadata lines are **automatically removed** from the document body (they only appear in the YAML)
- `Date` and `Description` metadata are optional

::: {.alert .alert-info}
**Rules for multi-line descriptions:**

The `Description` field can span multiple lines. To continue the description on a new line:

- Start the next line with `#` followed by **at least one space** and the continuation text
- Continuation lines are automatically concatenated with a space
- An empty line (or any other type of comment) ends the description

**Example:**

```r
# Title : Iris Data Analysis
#
# Author : Jane Doe
#
# Date : 2025-11-28
#
# Description : This analysis explores differences between iris species
# using various statistical methods and visualization techniques
# to identify patterns and correlations.
#

library(dplyr)
```

This will produce in the YAML:

```yaml
description: This analysis explores differences between iris species using various statistical methods and visualization techniques to identify patterns and correlations.
```
:::

### 1. Code Sections (Headers)

Use RStudio code section syntax to create headers at different levels. These sections MUST follow this exact pattern:

- **Level 2**: `## Title ####` (at least 4 `#` symbols at the end)
- **Level 3**: `### Title ====` (at least 4 `=` symbols at the end)
- **Level 4**: `#### Title ----` (at least 4 `-` symbols at the end)

::: {.alert .alert-warning}
<span style="color: red;">**Important rules:**

- There must be at least one space between the title text and the trailing symbols
- The trailing symbols must be at least 4 characters long
- You can use `#`, `=`, or `-` interchangeably (e.g., `## Title ====` works), but following the RStudio convention is recommended for consistency
- This follows the [RStudio code sections convention](https://docs.posit.co/ide/user/ide/guide/code/code-sections.html)</span>
:::

**Example:**

```r
## Data Loading ####

### Import CSV ====

#### Check Missing Values ----
```

### 2. Regular Comments (Text)

Single `#` comments become explanatory text in the Quarto document:

```r
# This is a regular comment
# It will appear as plain text in the output
# Use these to explain what your code does
```

**Rules:**

- Start with a single `#` followed by a space
- Multiple consecutive comment lines are grouped together (with no empty lines between them)
- Empty comment lines separate comment blocks
- **Markdown Tables**: You can include Markdown tables in comments. Consecutive comment lines will be preserved together, allowing proper table rendering

::: {.alert .alert-info}
**Important for Markdown Tables:** Table lines must be **isolated from other comments** with an empty line before and after the table. This ensures the table is treated as a separate block and will render correctly.
:::

**Markdown table example:**

```r
# Analysis results:

# | fruit  | price  |
# |--------|--------|
# | apple  | 2.05   |
# | pear   | 1.37   |
# | orange | 3.09   |
#
# The table above is properly isolated.
```

**Tip:** Use RStudio's [Comment/Uncomment shortcut](https://docs.posit.co/ide/user/ide/guide/productivity/text-editor.html#commentuncomment) (`Ctrl+Shift+C` on Windows/Linux or `Cmd+Shift+C` on Mac) to quickly add or remove comments from your code.

### 3. Code Lines

Any line that is NOT a comment becomes executable R code:

```r
library(dplyr)

iris |> 
  filter(Species == "setosa") |>
  summarize(mean_length = mean(Sepal.Length))
```

**Rules:**

- Consecutive code lines are grouped into a single code chunk
- Empty lines between code blocks are ignored
- Code blocks are separated by comments or section headers

**Important:** You can include comments WITHIN code blocks. These comments will be preserved inside the R code chunk:

```r
iris %>% 
  # Select a column
  select(Species)
```

This will render as:

````markdown
```{r, eval=FALSE}
#| eval: false
#| echo: true
iris %>% 
  # Select a column
  select(Species)
```
````

The difference:
- **Comments at the start of a line** (standalone) → become text OUTSIDE code blocks
- **Comments within code** (indented or part of a pipeline) → stay INSIDE code blocks

## Complete Example: example.R

Here's the complete example R script included in the package:

```r
library(dplyr)

## Title 2 ####

### Title 3 ====

# Start of statistical processing
# Counting the number of observations by species

iris |> 
  count(Species)

### Title 3 ====

# Filter the data.frame on Species "setosa"

iris |> 
  filter(Species == "setosa")

#### Title 4 ----

# Select column Species

iris %>% 
  # Select a column
  select(Species)
```

This script demonstrates:

1. **Code without section**: `library(dplyr)` becomes a code chunk
2. **Section headers**: `## Title 2 ####`, `### Title 3 ====`, `#### Title 4 ----`
3. **Standalone comments**: `# Select column Species` becomes readable text
4. **Inline comments**: `# Select a column` stays inside the code block
5. **Code blocks**: Separated by standalone comments or sections

## Customization Options

You can customize the output document with several parameters:

```{r eval=FALSE}
rtoqmd(
  input_file = "my_script.R",
  output_file = "my_document.qmd",
  title = "My Analysis Report",
  author = "Your Name",
  format = "html"
)
```

### Parameters

- `input_file`: Path to your R script
- `output_file`: Path for the output Quarto document (optional)
- `title`: Title for the document (default: "My title")
- `author`: Author name (default: "Your name")
- `format`: Output format - "html" or "pdf" (default: "html")
- `render`: Whether to render the .qmd to output file (default: TRUE)
- `open_html`: Whether to open the output file after rendering (default: FALSE)
- `code_fold`: Whether to fold code blocks by default (default: FALSE, HTML only)
- `number_sections`: Whether to number sections automatically (default: TRUE)

## Converting the Example

To convert the example script above:

```{r eval=FALSE}
# Get the example file path
example_file <- system.file("examples", "example.R", package = "quartify")

# Convert it
rtoqmd(
  input_file = example_file,
  output_file = "iris_analysis.qmd",
  title = "Iris Dataset Analysis",
  author = "Data Analyst"
)
```

### Generated Output

This produces the following Quarto document:

````markdown
---
title: "Iris Dataset Analysis"
author: "Data Analyst"
format:
  html:
    embed-resources: true
    code-fold: false
toc: true
toc-title: Sommaire
toc-depth: 4  
toc-location: left
execute: 
  eval: false
  echo: true
output: 
  html_document:
  number_sections: TRUE
  output-file: iris_analysis.html
---

```{r, eval=FALSE}
library(dplyr)
```

## Title 2

### Title 3

Start of statistical processing

Counting the number of observations by species

```{r}
#| eval: false
#| echo: true
iris |> 
  count(Species)
```

### Title 3

Filter the data.frame on Species "setosa"

```{r}
#| eval: false
#| echo: true
iris |> 
  filter(Species == "setosa")
```

#### Title 4

Select column Species

```{r}
#| eval: false
#| echo: true
iris %>% 
  # Select a column
  select(Species)
```
````

Notice how:

- Each code section becomes a proper markdown header
- Standalone comments become readable text paragraphs
- Inline comments (within code) are preserved in code blocks
- Code blocks are properly formatted with syntax highlighting
- The table of contents will show the hierarchical structure

**Important note about code chunks:**  
The generated code chunks use standard R chunks.  
The YAML header includes global `execute` options (`eval: false` and `echo: true`), which creates **non-executable** code blocks.   
Additionally, the HTML format uses `embed-resources: true` to create **self-contained HTML files** (see [Quarto documentation](https://quarto.org/docs/output-formats/html-basics.html#self-contained)).  
This is intentional - `quartify` is designed to create **static documentation** of your R script, not an executable notebook.  
The code is displayed with syntax highlighting for reading and documentation purposes, but won't be executed when rendering the Quarto document.  

This approach is ideal for: 

- Documenting existing scripts without modifying their execution
- Creating static references of your code
- Sharing code examples that readers can copy and run in their own environment
- Ensuring the documentation process doesn't interfere with your original script's behavior

## Rendering the Output

### Automatic Rendering (Recommended)

By default, `rtoqmd()` automatically renders your Quarto document to HTML:

```{r eval=FALSE}
# This will create both .qmd and .html files, then open the HTML
rtoqmd(example_file, "iris_analysis.qmd")
```

The function will:
1. Check if Quarto is installed
2. Render the .qmd file to HTML
3. Open the HTML file in your default browser (if `open_html = TRUE`)

If you don't want automatic rendering:

```{r eval=FALSE}
rtoqmd(example_file, "iris_analysis.qmd", render = FALSE)
```

### Manual Rendering

You can also render manually using Quarto:

```{bash eval=FALSE}
quarto render iris_analysis.qmd
```

Or from R:

```{r eval=FALSE}
quarto::quarto_render("iris_analysis.qmd")
```

**Note:** Quarto must be installed on your system. Download it from [quarto.org](https://quarto.org/docs/get-started/).

## Batch Converting Directories

To convert all R scripts in a directory (including subdirectories), use `rtoqmd_dir()`:

```{r eval=FALSE}
# Convert all R scripts in a directory
rtoqmd_dir("path/to/scripts")

# Convert and render to HTML
rtoqmd_dir("path/to/scripts", render = TRUE)

# With custom author and title prefix
rtoqmd_dir("path/to/scripts", 
           title_prefix = "Analysis: ",
           author = "Data Team")

# Exclude certain files (e.g., test files)
rtoqmd_dir("path/to/scripts", 
           exclude_pattern = "test_.*\\.R$")

# Non-recursive (current directory only)
rtoqmd_dir("path/to/scripts", recursive = FALSE)
```

This function:
- Recursively searches for all `.R` files in the directory
- Converts each file to `.qmd` in the same location
- Displays a conversion summary with success/failure counts
- Returns a data frame with results for each file

## CI/CD Integration

You can use `quartify` in your CI/CD pipelines to automatically generate documentation from your R scripts. This is useful for:

- Version controlling only R scripts (not generated documentation)
- Automatically generating `.qmd` and `.html` files in CI/CD
- Providing documentation artifacts for each commit/merge request

### GitHub Actions

Create `.github/workflows/generate-docs.yml`:

```yaml
name: Generate Documentation

on:
  push:
    branches: [main]
  pull_request:
    branches: [main]

jobs:
  generate-docs:
    runs-on: ubuntu-latest
    
    steps:
      - uses: actions/checkout@v4
      
      - uses: r-lib/actions/setup-r@v2
        with:
          r-version: '4.3.0'
      
      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2
      
      - name: Install R dependencies
        run: |
          install.packages(c("remotes", "dplyr"))
          remotes::install_github("ddotta/quartify")
        shell: Rscript {0}
      
      - name: Convert R scripts to Quarto
        run: |
          library(quartify)
          # Convert all R scripts in a directory
          rtoqmd_dir("scripts/", render = TRUE, author = "Data Team")
        shell: Rscript {0}
      
      - name: Upload artifacts
        uses: actions/upload-artifact@v4
        with:
          name: documentation
          path: |
            scripts/**/*.qmd
            scripts/**/*.html
```

### GitLab CI

Create `.gitlab-ci.yml`:

```yaml
generate-docs:
  image: rocker/r-ver:4.3.0
  
  before_script:
    # Install Quarto
    - apt-get update
    - apt-get install -y curl
    - curl -LO https://quarto.org/download/latest/quarto-linux-amd64.deb
    - dpkg -i quarto-linux-amd64.deb
    
    # Install R packages
    - R -e "install.packages(c('remotes', 'dplyr'))"
    - R -e "remotes::install_github('ddotta/quartify')"
  
  script:
    # Convert all R scripts to Quarto and render to HTML
    - R -e "quartify::rtoqmd_dir('scripts/', render = TRUE, author = 'Data Team')"
  
  artifacts:
    paths:
      - scripts/**/*.qmd
      - scripts/**/*.html
    expire_in: 30 days
  
  only:
    - main
    - merge_requests
```

### Single File Conversion in CI/CD

For converting a single R script:

```yaml
# GitHub Actions
- name: Convert single script
  run: |
    library(quartify)
    rtoqmd("analysis.R", "analysis.qmd", 
           title = "My Analysis",
           author = "Data Team",
           render = TRUE)
  shell: Rscript {0}

# GitLab CI
script:
  - R -e "quartify::rtoqmd('analysis.R', 'analysis.qmd', title = 'My Analysis', author = 'Data Team', render = TRUE)"
```

### Best Practices for CI/CD

1. **Add to .gitignore**: Exclude generated files from version control
   ```
   # .gitignore
   *.qmd
   *.html
   *_files/
   ```

2. **Artifact retention**: Set appropriate expiration for artifacts
   - Short-lived (7 days) for feature branches
   - Long-lived (90+ days) for main branch

3. **Conditional execution**: Only generate docs when R scripts change
   ```yaml
   # GitHub Actions
   on:
     push:
       paths:
         - 'scripts/**/*.R'
   ```

4. **Error handling**: Use `render = FALSE` to debug conversion issues
   ```r
   # First convert without rendering
   rtoqmd_dir("scripts/", render = FALSE)
   # Then render separately if needed
   ```

## Use Cases

`quartify` is particularly useful for:

1. **Documentation**: Transform working scripts into documentation
2. **Code review**: Present code in a more accessible format
3. **Batch conversion**: Automatically convert all scripts in a project

## Callouts

Callouts are special blocks that highlight important information in your Quarto documents. `quartify` supports automatic conversion of callout comments.

### Syntax

In your R script, use this syntax:

```r
# callout-tip - Tip with Title
# This is the content of the note.
# It can span multiple lines.

# Code or empty line ends the callout
x <- 1
```

This converts to:

```markdown
::: {.callout-tip title="Tip with Title"}
This is the content of the note.
It can span multiple lines.
:::
```

### Callout Types

Five callout types are supported:

- `# callout-note` - Blue information box
- `# callout-tip` - Green tip box
- `# callout-warning` - Orange warning box
- `# callout-caution` - Red caution box
- `# callout-important` - Red important box

### With or Without Title

**With title:**
```r
# callout-tip - Useful Tip
# Use callouts to highlight key information.
```

**Without title:**
```r
# callout-tip
# Use callouts to highlight key information.
```

### Examples

```r
# callout-note - Data Requirements
# The input data must contain columns: Species, Sepal.Length, Petal.Length

# callout-warning
# This operation may take several minutes with large datasets

# callout-tip - Performance Optimization
# Use data.table for datasets larger than 1GB

library(dplyr)
iris %>% head()
```

### Rules

- Callout starts with `# callout-TYPE` (optionally followed by ` - Title`)
- All subsequent comment lines are part of the callout content
- Callout ends when encountering:
  - An empty line
  - A code line
  - Another section or callout
- Title is optional (omit the ` - Title` part)

::: {.alert .alert-info}
**Tip:** Create RStudio snippets for callouts to speed up your workflow:

```
snippet callout
	# callout-${1:note} - ${2:Title}
	# ${0}
```

Type `callout` followed by `Tab` to insert the template.
:::

## Comment Rules Summary

| Type | Syntax | Result | Example |
|------|--------|--------|---------|
| **Level 2 Header** | `## Title ####` | Markdown `## Title` | `## Data Analysis ####` |
| **Level 3 Header** | `### Title ====` | Markdown `### Title` | `### Preprocessing ====` |
| **Level 4 Header** | `#### Title ----` | Markdown `#### Title` | `#### Remove NA ----` |
| **Standalone Comment** | `# Text` | Plain text paragraph | `# This filters the data` |
| **Callout** | `# callout-TYPE - Title` | Quarto callout block | `# callout-note - Important` |
| **Code** | No `#` prefix | R code chunk | `iris %>% filter(...)` |
| **Inline Comment** | `  # Text` in code | Stays in code chunk | `iris %>% # comment` |

**Critical rules to avoid errors:**

1. **Spacing**: Always include a space after `#` for comments and section headers
2. **Trailing symbols**: Section headers MUST have at least 4 trailing symbols
3. **Symbol flexibility**: You can use `#`, `=`, or `-` interchangeably for trailing symbols, but following the RStudio convention is recommended
4. **Roxygen comments**: Lines starting with `#'` are ignored (for package development)
5. **Comment context**: Comments at the start of a line become text; comments within code stay in code blocks
6. **Keyboard shortcut**: Use `Ctrl+Shift+C` (Windows/Linux) or `Cmd+Shift+C` (Mac) to comment/uncomment lines quickly

## Tips and Best Practices

1. **Start with structure**: Define your section headers first to create the document outline
2. **Use consistent levels**: Follow a logical hierarchy (2 → 3 → 4, don't skip levels)
3. **Add explanatory text**: Use regular comments to explain what your code does
4. **Group related code**: Keep related operations together; they'll be grouped into the same code block
5. **Test incrementally**: Start with a small script to see how the conversion works
6. **Navigate easily**: In RStudio, use the document outline (Ctrl+Shift+O) to see your structure
7. **Comment liberally**: More comments = better documentation in the final Quarto document

## Conclusion

`quartify` makes it easy to transform your R scripts into professional Quarto documents without manual reformatting. By following RStudio code section conventions, you can automatically generate well-structured, reproducible documentation from your existing code with proper navigation hierarchy.
