---
title: "Getting Started with quartify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with quartify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`quartify` is an R package that automatically converts R scripts into Quarto markdown documents (.qmd). This vignette will guide you through the basic usage and features of the package.

## Installation

You can install the development version of quartify from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ddotta/quartify")
```

## Basic Usage

The main function of the package is `rtoqmd()`. Here's a simple example:

```{r eval=FALSE}
library(quartify)

# Convert an R script to a Quarto document
rtoqmd("my_script.R", "my_document.qmd")
```

## Structuring Your R Script

For optimal conversion, structure your R script using special comment syntax:

### Headers

Use `# ##` for level 2 headers, `# ###` for level 3, and so on up to level 6:

```r
# ## Main Section

# ### Subsection

# #### Sub-subsection
```

### Comments

Regular comments (single `#`) become explanatory text in the output:

```r
# This comment will appear as plain text
# in the Quarto document
```

### Code

Any non-commented code is automatically grouped into executable code blocks:

```r
library(dplyr)

iris |> 
  filter(Species == "setosa") |>
  summarize(mean_length = mean(Sepal.Length))
```

## Customization Options

You can customize the output document with several parameters:

```{r eval=FALSE}
rtoqmd(
  input_file = "my_script.R",
  output_file = "my_document.qmd",
  title = "My Analysis Report",
  author = "Your Name",
  format = "html"
)
```

### Parameters

- `input_file`: Path to your R script
- `output_file`: Path for the output Quarto document (optional)
- `title`: Title for the document (default: "My title")
- `author`: Author name (default: "Damien Dotta")
- `format`: Output format (default: "html")

## Complete Example

Let's look at a complete example using the included sample file:

```{r eval=FALSE}
# Get the example file path
example_file <- system.file("examples", "example.R", package = "quartify")

# Convert it
rtoqmd(
  input_file = example_file,
  output_file = "iris_analysis.qmd",
  title = "Iris Dataset Analysis",
  author = "Data Analyst"
)
```

This will create a Quarto document with:

- A YAML header with title, author, and format information
- A table of contents
- Properly formatted headers
- Explanatory text from comments
- Executable R code blocks

## Rendering the Output

Once you have your `.qmd` file, you can render it to HTML using Quarto:

```{bash eval=FALSE}
quarto render iris_analysis.qmd
```

Or from R:

```{r eval=FALSE}
quarto::quarto_render("iris_analysis.qmd")
```

## Use Cases

`quartify` is particularly useful for:

1. **Documentation**: Transform working scripts into professional documentation
2. **Sharing analyses**: Create readable reports from existing code
3. **Reproducible research**: Combine code and narrative seamlessly
4. **Code review**: Present code in a more accessible format

## Tips and Best Practices

1. **Structure your comments**: Use header comments (`# ##`, `# ###`) to organize your analysis into logical sections
2. **Add explanatory text**: Use regular comments to explain what your code does
3. **Group related code**: Keep related operations together; they'll be grouped into the same code block
4. **Test incrementally**: Start with a small script to see how the conversion works

## Conclusion

`quartify` makes it easy to transform your R scripts into professional Quarto documents without manual reformatting. By following simple commenting conventions, you can automatically generate well-structured, reproducible documentation from your existing code.
