---
title: "Getting Started with quartify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with quartify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {.alert .alert-info}
🇫🇷 **[Version française disponible](getting-started_FR.html)** / **French version available**
:::

## Introduction

`quartify` is an R package that automatically converts R scripts into Quarto markdown documents (.qmd). The package recognizes RStudio code sections to create properly structured documents with navigation. This vignette will guide you through the basic usage and features of the package.

### Comparison with knitr::spin()

If you're familiar with [`knitr::spin()`](https://yihui.org/knitr/demo/stitch/), you might wonder how `quartify` differs:

| Feature | knitr::spin() | quartify |
|---------|---------------|----------|
| **Output Format** | R Markdown (.Rmd) | Quarto (.qmd) |
| **Text Syntax** | `#'` for markdown | `#` (natural R comments) |
| **Chunk Options** | `#+` prefix | RStudio code sections |
| **Modern Features** | Limited to R Markdown | Full Quarto capabilities |
| **Callouts** | Manual HTML/CSS | Native Quarto callouts |
| **Diagrams** | External tools | Native Mermaid support |
| **Themes** | Limited R Markdown themes | 25+ modern Quarto themes |

**Why choose quartify?**

1. **Modern Quarto Ecosystem**: Benefit from Quarto's next-generation features (callouts, tabsets, cross-references, Observable JS, etc.)
2. **Cleaner R Scripts**: Use natural R commenting without special `#'` or `#+` syntax - your scripts remain readable as standalone R code
3. **Advanced Outputs**: Generate websites, books, presentations, and scientific articles with the same workflow
4. **Better Theming**: Access to modern, responsive HTML themes with consistent styling
5. **Future-Proof**: Quarto is actively developed by Posit as the successor to R Markdown

S## Installation

You can install the development version of quartify from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ddotta/quartify")
```

## Basic Usage

The main function of the package is `rtoqmd()`. Here's a simple example:

```{r eval=FALSE}
library(quartify)

# Convert an R script to a Quarto document and render to HTML
rtoqmd("my_script.R", "my_document.qmd")

# Convert only, without rendering to HTML
rtoqmd("my_script.R", "my_document.qmd", render = FALSE)
```

By default, `rtoqmd()` will:
1. Create the .qmd file
2. Render it to HTML using Quarto
3. Open the HTML file in your default browser

### Output File Locations

Understanding where files are created:

- **If you specify an `output_file` path**: The .qmd file is created at that exact location, and the .html file (if rendered) is created in the same directory with the same base name.

```{r eval=FALSE}
# Creates: /path/to/my_analysis.qmd and /path/to/my_analysis.html
rtoqmd("script.R", "/path/to/my_analysis.qmd")
```

- **If you don't specify `output_file`**: The .qmd file is created in the same directory as your input R script, with the `.R` extension replaced by `.qmd`.

```{r eval=FALSE}
# If script.R is in /home/user/scripts/
# Creates: /home/user/scripts/script.qmd and /home/user/scripts/script.html
rtoqmd("/home/user/scripts/script.R")
```

- **Relative paths**: If you use relative paths, files are created relative to your current working directory (check with `getwd()`).

```{r eval=FALSE}
# Creates files in your current working directory
rtoqmd("script.R", "output.qmd")
```

## Using the RStudio Addin

`quartify` provides an interactive RStudio addin for easy conversion without writing code. This is perfect for quick conversions while working in RStudio.

### Accessing the Addin

To use the addin:

1. **From RStudio menu**: Go to **Addins** > **"Convert R to Quarto (rtoqmd)"**
2. The interactive interface will open in a new window

::: {.alert .alert-info}
**Tip:** You can also bind a keyboard shortcut to the addin for even faster access. Go to **Tools** > **Modify Keyboard Shortcuts** and search for "quartify" to assign your preferred shortcut.
:::

### Interface Overview

The addin features a modern, intuitive interface with:

- **Blue title bar** with language switcher (EN/FR flags)
- **GENERATE button** prominently placed at the top below the title bar, next to the quartify logo
- **Mode selector** to choose between single file or directory conversion
- **File/directory browser** for easy selection
- **Real-time validation** and error messages

### Single File Mode

Perfect for converting one R script at a time:

1. **Select mode**: Choose "Single File Mode"
2. **Browse for input**: Click "Browse" to select your R script
3. **Set output location**: Optionally specify where to save the .qmd file (defaults to same directory as input)
4. **Configure options**:
   - **Render to HTML**: Check to automatically generate HTML output
   - **Open HTML**: Check to open the result in your browser
5. **Click GENERATE**: The blue GENERATE button at the top starts the conversion

The addin will show success/error messages and provide clickable links to output files.

### Directory Mode

Convert all R scripts in a directory at once:

1. **Select mode**: Choose "Directory Mode"
2. **Browse for input directory**: Select the folder containing your R scripts
3. **Configure options**:
   - **Recursive**: Check to include subdirectories
   - **Render to HTML**: Generate HTML for all files
   - **Create Book**: Automatically create a `_quarto.yml` file to combine all documents into a Quarto book
4. **Output directory** (Optional): Choose where to save converted files (defaults to same location as input files)
5. **Click GENERATE**: Start batch conversion

::: {.callout-note}
**Create Book Feature**: When enabled, this option creates a `_quarto.yml` configuration file in the output directory, allowing you to render all converted documents as a unified Quarto book with automatic navigation and consistent styling. This is ideal for project documentation or analysis collections.
:::

### Advanced Features

The addin includes several quality-of-life improvements:

- **Volume management**: Easily navigate across different drives and network locations
- **Path validation**: Real-time checks for valid file/directory paths
- **Bilingual interface**: Switch between English and French instantly
- **Persistent settings**: Your last mode selection is remembered
- **Detailed feedback**: Clear success/error messages with file counts and locations

### When to Use the Addin

The RStudio addin is ideal for:

- **Interactive exploration**: Testing conversion on sample scripts
- **One-off conversions**: Quick document generation without scripting
- **Visual feedback**: Seeing results immediately with the GUI
- **Learning the package**: Understanding options before automating
- **Teaching**: Demonstrating quartify features to colleagues

For batch processing or CI/CD pipelines, consider using the programmatic functions (`rtoqmd()`, `rtoqmd_dir()`) instead.

## Using the Standalone Shiny App

For users working outside RStudio or wanting to share a conversion tool with their team, `quartify` provides a standalone Shiny application.

### Launching the App

Simply run:

```{r eval=FALSE}
library(quartify)
quartify_app()
```

This opens the same intuitive interface as the RStudio addin, but in a standard browser window. The app can run on any system with R installed, making it perfect for:

- **Non-RStudio environments**: Use with VS Code, Jupyter, or other IDEs
- **Shared team tools**: Deploy on RStudio Connect or Shiny Server for team access
- **Demonstrations**: Show quartify features without requiring RStudio
- **Remote work**: Access from any browser if hosted on a server

### Features

The standalone app includes all the same features as the RStudio addin:

- **Single File Mode**: Convert individual R scripts with full control over output location and rendering options
- **Directory Mode**: Batch convert entire folders with recursive directory scanning
- **Create Book**: Generate `_quarto.yml` configuration for Quarto book projects
- **Bilingual Interface**: Switch between English and French with flag buttons
- **Modern UI**: Blue title bar, prominent GENERATE button, and intuitive layout

### Comparison with Addin

| Feature | RStudio Addin | Standalone App |
|---------|---------------|----------------|
| **Requires RStudio** | Yes | No |
| **Launch Method** | Addins menu | `quartify_app()` |
| **Interface** | Gadget window | Browser window |
| **File Browser** | Native file system | Native file system |
| **All Features** | ✓ | ✓ |
| **Shareable** | No | Yes (can deploy) |

### Deployment Options

The standalone app can be deployed for team access:

**Local Network:**
```{r eval=FALSE}
# Run on a specific port accessible to your network
shiny::runApp(system.file("shiny", "quartify_app", package = "quartify"), 
              host = "0.0.0.0", port = 3838)
```

**RStudio Connect / Shiny Server:**
Deploy as a standard Shiny application for enterprise-wide access.

### Web Version

For environments where file system access is restricted (cloud deployments, sandboxed environments), use the web version:

```{r eval=FALSE}
quartify_app_web()
```

This version features:

- **File upload**: Upload R scripts directly from your computer (single or multiple files)
- **File download**: Download generated files as .zip archives with organized structure
- **ZIP structure**: 
  - `qmd/` folder with all Quarto markdown files and configuration
  - `html/` folder with complete rendered book (including `index.html` and all resources)
- **Batch conversion**: Upload multiple files to create a complete Quarto book
- **Offline viewing**: Downloaded ZIP contains everything needed to view the book offline
- **Same conversion logic**: All quartify features available

Perfect for cloud deployments where direct file system access isn't available. The downloaded ZIP is ready to extract and view immediately in any browser.

## Structuring Your R Script

For optimal conversion, you need to follow specific commenting rules in your R script. `quartify` recognizes three types of lines:

### 0. Document Metadata (Optional)

You can define your document's metadata directly in your R script using special comments at the beginning of the file. These metadata will appear in the YAML header of the generated Quarto document.

**Recognized metadata:**

- **Title**: `# Title : My title` or `# Titre : Mon titre`
- **Author**: `# Author : My name` or `# Auteur : Mon nom`
- **Date**: `# Date : 2025-11-28`
- **Description**: `# Description : Description of your script`

::: {.alert .alert-success}
**Tip - RStudio Snippets:** quartify provides a function to automatically install useful snippets including `header`, `callout`, `mermaid`, and `tabset`:

```{r eval=FALSE}
# Install quartify snippets
install_quartify_snippets()
```

After installation and restarting RStudio, type the snippet name (e.g., `header`) followed by `Tab` in your R script to automatically insert the template.

You can also manually add snippets via Tools > Edit Code Snippets > R. See `?install_quartify_snippets` for more details.
:::

**Complete example with metadata:**

```r
# Title : Iris Data Analysis
#
# Author : Jane Doe
#
# Date : 2025-11-28
#
# Description : Explore differences between iris species
#

library(dplyr)

## Descriptive Analysis ####

# Calculate statistics by species

iris |> 
  group_by(Species) |>
  summarize(mean_sepal = mean(Sepal.Length))
```

**Behavior:**

- If metadata is found in the script, it **overrides** the `rtoqmd()` function parameters
- If no metadata is found, the function's `title` and `author` parameters are used
- Metadata lines are **automatically removed** from the document body (they only appear in the YAML)
- `Date` and `Description` metadata are optional

::: {.alert .alert-info}
**Rules for multi-line descriptions:**

The `Description` field can span multiple lines. To continue the description on a new line:

- Start the next line with `#` followed by **at least one space** and the continuation text
- Continuation lines are automatically concatenated with a space
- An empty line (or any other type of comment) ends the description

**Example:**

```r
# Title : Iris Data Analysis
#
# Author : Jane Doe
#
# Date : 2025-11-28
#
# Description : This analysis explores differences between iris species
# using various statistical methods and visualization techniques
# to identify patterns and correlations.
#

library(dplyr)
```

This will produce in the YAML:

```yaml
description: This analysis explores differences between iris species using various statistical methods and visualization techniques to identify patterns and correlations.
```
:::

### 1. Code Sections (Headers)

Use RStudio code section syntax to create headers at different levels. These sections MUST follow this exact pattern:

- **Level 2**: `## Title ####` (at least 4 `#` symbols at the end)
- **Level 3**: `### Title ====` (at least 4 `=` symbols at the end)
- **Level 4**: `#### Title ----` (at least 4 `-` symbols at the end)

::: {.alert .alert-warning}
<span style="color: red;">**Important rules:**

- There must be at least one space between the title text and the trailing symbols
- The trailing symbols must be at least 4 characters long
- You can use `#`, `=`, or `-` interchangeably (e.g., `## Title ====` works), but following the RStudio convention is recommended for consistency
- This follows the [RStudio code sections convention](https://docs.posit.co/ide/user/ide/guide/code/code-sections.html)</span>
:::

**Example:**

```r
## Data Loading ####

### Import CSV ====

#### Check Missing Values ----
```

### 2. Regular Comments (Text)

Single `#` comments become explanatory text in the Quarto document:

```r
# This is a regular comment
# It will appear as plain text in the output
# Use these to explain what your code does
```

**Rules:**

- Start with a single `#` **followed by a space** (no space before the `#`)
- Multiple consecutive comment lines are grouped together (with no empty lines between them)
- Empty comment lines separate comment blocks
- **Markdown Tables**: You can include Markdown tables in comments. Consecutive comment lines will be preserved together, allowing proper table rendering

::: {.alert .alert-warning}
**⚠️ Important for text conversion:**

For a comment to be correctly converted to Quarto text, the line must start with `#` **without any leading space**. If you have indentation spaces before the `#`, the comment will be ignored and remain in the code chunk.

**Correct example:**
```r
# This will be converted to text
result <- mean(x)
```

**Incorrect example (comment stays in code) :**
```r
if (condition) {
  # This will remain an R comment in the chunk
  result <- mean(x)
}
```
:::

::: {.callout-tip title="Tip: Split a long code chunk"}
You can use this rule to **split a long code chunk into multiple parts**. By inserting a comment **at the start of a line** (no space before `#`) between two code blocks, this comment will be converted to text and naturally create two separate chunks:

```r
data <- read.csv("file.csv")
data_clean <- na.omit(data)

# This comment splits the chunk into two parts

result <- mean(data_clean$value)
plot(result)
```

This will create two separate chunks instead of one large chunk.
:::

::: {.alert .alert-info}
**Important for Markdown Tables:** Table lines must be **isolated from other comments** with an empty line before and after the table. This ensures the table is treated as a separate block and will render correctly.
:::

**Markdown table example:**

```r
# Analysis results:

# | fruit  | price  |
# |--------|--------|
# | apple  | 2.05   |
# | pear   | 1.37   |
# | orange | 3.09   |
#
# The table above is properly isolated.
```

**Tip:** Use RStudio's [Comment/Uncomment shortcut](https://docs.posit.co/ide/user/ide/guide/productivity/text-editor.html#commentuncomment) (`Ctrl+Shift+C` on Windows/Linux or `Cmd+Shift+C` on Mac) to quickly add or remove comments from your code.

### 3. Code Lines

Any line that is NOT a comment becomes executable R code:

```r
library(dplyr)

iris |> 
  filter(Species == "setosa") |>
  summarize(mean_length = mean(Sepal.Length))
```

**Rules:**

- Consecutive code lines are grouped into a single code chunk
- Empty lines between code blocks are ignored
- Code blocks are separated by comments or section headers

**Important:** You can include comments WITHIN code blocks. These comments will be preserved inside the R code chunk:

```r
iris %>% 
  # Select a column
  select(Species)
```

This will render as:

````markdown
```{r, eval=FALSE}
#| eval: false
#| echo: true
iris %>% 
  # Select a column
  select(Species)
```
````

The difference:  
- **Comments at the start of a line** (standalone) → become text OUTSIDE code blocks  
- **Comments within code** (indented or part of a pipeline) → stay INSIDE code blocks

## Complete Example: example.R

Here's the complete example R script included in the package:

```r
library(dplyr)

## Title 2 ####

### Title 3 ====

# Start of statistical processing
# Counting the number of observations by species

iris |> 
  count(Species)

### Title 3 ====

# Filter the data.frame on Species "setosa"

iris |> 
  filter(Species == "setosa")

#### Title 4 ----

# Select column Species

iris %>% 
  # Select a column
  select(Species)
```

This script demonstrates:

1. **Code without section**: `library(dplyr)` becomes a code chunk
2. **Section headers**: `## Title 2 ####`, `### Title 3 ====`, `#### Title 4 ----`
3. **Standalone comments**: `# Select column Species` becomes readable text
4. **Inline comments**: `# Select a column` stays inside the code block
5. **Code blocks**: Separated by standalone comments or sections

## Customization Options

You can customize the output document with several parameters:

```{r eval=FALSE}
rtoqmd(
  input_file = "my_script.R",
  output_file = "my_document.qmd",
  title = "My Analysis Report",
  author = "Your Name",
  format = "html"
)
```

### Parameters

- `input_file`: Path to your R script
- `output_file`: Path for the output Quarto document (optional)
- `title`: Title for the document (default: "My title")
- `author`: Author name (default: "Your name")
- `format`: Output format - always "html" (parameter kept for backward compatibility)
- `render`: Whether to render the .qmd to output file (default: TRUE)
- `open_html`: Whether to open the output file after rendering (default: FALSE)
- `code_fold`: Whether to fold code blocks by default (default: FALSE, HTML only)
- `number_sections`: Whether to number sections automatically (default: TRUE)

## Converting the Example

To convert the example script above:

```{r eval=FALSE}
# Get the example file path
example_file <- system.file("examples", "example.R", package = "quartify")

# Convert it
rtoqmd(
  input_file = example_file,
  output_file = "iris_analysis.qmd",
  title = "Iris Dataset Analysis",
  author = "Data Analyst"
)
```

### Generated Output

This produces the following Quarto document:

````markdown
---
title: "Iris Dataset Analysis"
author: "Data Analyst"
format:
  html:
    embed-resources: true
    code-fold: false
toc: true
toc-title: Sommaire
toc-depth: 4  
toc-location: left
execute: 
  eval: false
  echo: true
output: 
  html_document:
  number_sections: TRUE
  output-file: iris_analysis.html
---

```{r, eval=FALSE}
library(dplyr)
```

## Title 2

### Title 3

Start of statistical processing

Counting the number of observations by species

```{r}
#| eval: false
#| echo: true
iris |> 
  count(Species)
```

### Title 3

Filter the data.frame on Species "setosa"

```{r}
#| eval: false
#| echo: true
iris |> 
  filter(Species == "setosa")
```

#### Title 4

Select column Species

```{r}
#| eval: false
#| echo: true
iris %>% 
  # Select a column
  select(Species)
```
````

Notice how:

- Each code section becomes a proper markdown header
- Standalone comments become readable text paragraphs
- Inline comments (within code) are preserved in code blocks
- Code blocks are properly formatted with syntax highlighting
- The table of contents will show the hierarchical structure

**Important note about code chunks:**  
The generated code chunks use standard R chunks.  
The YAML header includes global `execute` options (`eval: false` and `echo: true`), which creates **non-executable** code blocks.   
Additionally, the HTML format uses `embed-resources: true` to create **self-contained HTML files** (see [Quarto documentation](https://quarto.org/docs/output-formats/html-basics.html#self-contained)).  
This is intentional - `quartify` is designed to create **static documentation** of your R script, not an executable notebook.  
The code is displayed with syntax highlighting for reading and documentation purposes, but won't be executed when rendering the Quarto document.  

This approach is ideal for: 

- Documenting existing scripts without modifying their execution
- Creating static references of your code
- Sharing code examples that readers can copy and run in their own environment
- Ensuring the documentation process doesn't interfere with your original script's behavior

## Rendering the Output

### Automatic Rendering (Recommended)

By default, `rtoqmd()` automatically renders your Quarto document to HTML:

```{r eval=FALSE}
# This will create both .qmd and .html files, then open the HTML
rtoqmd(example_file, "iris_analysis.qmd")
```

The function will:
1. Check if Quarto is installed
2. Render the .qmd file to HTML
3. Open the HTML file in your default browser (if `open_html = TRUE`)

If you don't want automatic rendering:

```{r eval=FALSE}
rtoqmd(example_file, "iris_analysis.qmd", render = FALSE)
```

### Manual Rendering

You can also render manually using Quarto:

```{bash eval=FALSE}
quarto render iris_analysis.qmd
```

Or from R:

```{r eval=FALSE}
quarto::quarto_render("iris_analysis.qmd")
```

**Note:** Quarto must be installed on your system. Download it from [quarto.org](https://quarto.org/docs/get-started/).

## Batch Converting Directories

`quartify` v0.0.7 introduces powerful batch conversion capabilities through multiple interfaces. Choose the method that best fits your workflow.

### Using the RStudio Addin or Standalone App

The most intuitive way to batch convert files is through the **Directory Mode** in either the RStudio addin or standalone app:

```{r eval=FALSE}
# Launch the standalone app
quartify_app()

# Or use the RStudio addin: Addins > "Convert R to Quarto (rtoqmd)"
```

**Directory Mode features (v0.0.7):**

1. **Browse for directory**: Select the folder containing your R scripts
2. **Recursive option**: Include or exclude subdirectories
3. **Output directory**: Optionally specify a different location for converted files (new in v0.0.7!)
4. **Create Book checkbox**: Automatically generate `_quarto.yml` to combine all documents into a Quarto book (new in v0.0.7!)
5. **Render option**: Choose whether to generate HTML files for all scripts
6. **One-click conversion**: Click the GENERATE button at the top to convert all files

::: {.callout-tip}
**New in v0.0.7**: The output directory feature allows you to keep your source R scripts separate from generated .qmd files. Combined with the "Create Book" option, you can instantly create a complete Quarto book project from a collection of R scripts.
:::

**Workflow example:**

- Input directory: `~/my-project/scripts/` (contains 10 R scripts)
- Output directory: `~/my-project/docs/` (where you want .qmd files)
- Enable "Create Book": Generates `~/my-project/docs/_quarto.yml`
- Result: Complete Quarto book ready to render with `quarto render`

### Using the Web Interface

For environments without direct file system access:

```{r eval=FALSE}
quartify_app_web()
```

In batch mode, you can either:
- **Upload multiple R files** at once
- **Select a directory** containing your R scripts

The interface will convert all files and provide a downloadable .zip archive.

### Using R Functions

To convert all R scripts in a directory (including subdirectories) programmatically, use `rtoqmd_dir()`:

```{r eval=FALSE}
# Convert all R scripts in a directory
rtoqmd_dir("path/to/scripts")

# Convert and render to HTML
rtoqmd_dir("path/to/scripts", render = TRUE)

# With custom author and title prefix
rtoqmd_dir("path/to/scripts", 
           title_prefix = "Analysis: ",
           author = "Data Team")

# Exclude certain files (e.g., test files)
rtoqmd_dir("path/to/scripts", 
           exclude_pattern = "test_.*\\.R$")

# Non-recursive (current directory only)
rtoqmd_dir("path/to/scripts", recursive = FALSE)
```

This function:
- Recursively searches for all `.R` files in the directory
- Converts each file to `.qmd` in the same location
- Displays a conversion summary with success/failure counts
- Returns a data frame with results for each file

## CI/CD Integration

You can use `quartify` in your CI/CD pipelines to automatically generate documentation from your R scripts. This is useful for:

- Version controlling only R scripts (not generated documentation)
- Automatically generating `.qmd` and `.html` files in CI/CD
- Providing documentation artifacts for each commit/merge request

### GitHub Actions

Create `.github/workflows/generate-docs.yml`:

```yaml
name: Generate Documentation

on:
  push:
    branches: [main]
  pull_request:
    branches: [main]

jobs:
  generate-docs:
    runs-on: ubuntu-latest
    
    steps:
      - uses: actions/checkout@v4
      
      - uses: r-lib/actions/setup-r@v2
        with:
          r-version: '4.5.1'
      
      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2
      
      - name: Install R dependencies
        run: |
          install.packages(c("remotes", "dplyr"))
          remotes::install_github("ddotta/quartify")
        shell: Rscript {0}
      
      - name: Convert R scripts to Quarto
        run: |
          library(quartify)
          # Convert all R scripts in a directory
          rtoqmd_dir("scripts/", render = TRUE, author = "Data Team")
        shell: Rscript {0}
      
      - name: Upload artifacts
        uses: actions/upload-artifact@v4
        with:
          name: documentation
          path: |
            scripts/**/*.qmd
            scripts/**/*.html
```

### GitLab CI

Create `.gitlab-ci.yml`:

```yaml
generate-docs:
  image: rocker/r-ver:4.5.1
  
  before_script:
    # Install Quarto
    - apt-get update
    - apt-get install -y curl
    - curl -LO https://quarto.org/download/latest/quarto-linux-amd64.deb
    - dpkg -i quarto-linux-amd64.deb
    
    # Install R packages
    - R -e "install.packages(c('remotes', 'dplyr'))"
    - R -e "remotes::install_github('ddotta/quartify')"
  
  script:
    # Convert all R scripts to Quarto and render to HTML
    - R -e "quartify::rtoqmd_dir('scripts/', render = TRUE, author = 'Data Team')"
  
  artifacts:
    paths:
      - scripts/**/*.qmd
      - scripts/**/*.html
    expire_in: 30 days
  
  only:
    - main
    - merge_requests
```

### Single File Conversion in CI/CD

For converting a single R script:

```yaml
# GitHub Actions
- name: Convert single script
  run: |
    library(quartify)
    rtoqmd("analysis.R", "analysis.qmd", 
           title = "My Analysis",
           author = "Data Team",
           render = TRUE)
  shell: Rscript {0}

# GitLab CI
script:
  - R -e "quartify::rtoqmd('analysis.R', 'analysis.qmd', title = 'My Analysis', author = 'Data Team', render = TRUE)"
```

### Best Practices for CI/CD

1. **Add to .gitignore**: Exclude generated files from version control
   ```
   # .gitignore
   *.qmd
   *.html
   *_files/
   ```

2. **Artifact retention**: Set appropriate expiration for artifacts
   - Short-lived (7 days) for feature branches
   - Long-lived (90+ days) for main branch

3. **Conditional execution**: Only generate docs when R scripts change
   ```yaml
   # GitHub Actions
   on:
     push:
       paths:
         - 'scripts/**/*.R'
   ```

4. **Error handling**: Use `render = FALSE` to debug conversion issues
   ```r
   # First convert without rendering
   rtoqmd_dir("scripts/", render = FALSE)
   # Then render separately if needed
   ```

## Use Cases

`quartify` is particularly useful for:

1. **Documentation**: Transform working scripts into documentation
2. **Code review**: Present code in a more accessible format
3. **Batch conversion**: Automatically convert all scripts in a project

## Callouts

Callouts are special blocks that highlight important information in your Quarto documents. `quartify` supports automatic conversion of callout comments.

### Syntax

In your R script, use this syntax:

```r
# callout-tip - Tip with Title
# This is the content of the note.
# It can span multiple lines.

# Code or empty line ends the callout
x <- 1
```

This converts to:

```markdown
::: {.callout-tip title="Tip with Title"}
This is the content of the note.
It can span multiple lines.
:::
```

### Callout Types

Five callout types are supported:

- `# callout-note` - Blue information box
- `# callout-tip` - Green tip box
- `# callout-warning` - Orange warning box
- `# callout-caution` - Red caution box
- `# callout-important` - Red important box

### With or Without Title

**With title:**
```r
# callout-tip - Useful Tip
# Use callouts to highlight key information.
```

**Without title:**
```r
# callout-tip
# Use callouts to highlight key information.
```

### Examples

```r
# callout-note - Data Requirements
# The input data must contain columns: Species, Sepal.Length, Petal.Length

# callout-warning
# This operation may take several minutes with large datasets

# callout-tip - Performance Optimization
# Use data.table for datasets larger than 1GB

library(dplyr)
iris %>% head()
```

### Rules

- Callout starts with `# callout-TYPE` (optionally followed by ` - Title`)
- All subsequent comment lines are part of the callout content
- Callout ends when encountering:
  - An empty line
  - A code line
  - Another section or callout
- Title is optional (omit the ` - Title` part)

::: {.alert .alert-info}
**Tip:** Use the `install_quartify_snippets()` function to add a `callout` snippet to your RStudio. Type `callout` followed by `Tab` to quickly insert the callout template.
:::

## Mermaid Diagrams

Create flowcharts, sequence diagrams, and other visualizations using Mermaid syntax. Mermaid diagrams are rendered directly in the HTML output.

### Syntax

In your R script, start with `#| mermaid`, add options with `#|`, then the diagram content:

```r
#| mermaid
#| eval: true
flowchart TD
    A[Start] --> B[Process Data]
    B --> C{Valid?}
    C -->|Yes| D[Save Results]
    C -->|No| E[Error]
    D --> F[End]
    E --> F
```

This converts to a proper Quarto Mermaid chunk that renders as an interactive diagram in HTML.

### Diagram Types

Mermaid supports many diagram types:

**Flowcharts:**
```r
#| mermaid
#| eval: true
flowchart LR
    A[Input] --> B[Process]
    B --> C[Output]
```

**Sequence Diagrams:**
```r
#| mermaid
#| eval: true
sequenceDiagram
    participant User
    participant API
    participant Database
    User->>API: Request data
    API->>Database: Query
    Database-->>API: Results
    API-->>User: Response
```

**Class Diagrams:**
```r
#| mermaid
#| eval: true
classDiagram
    class Animal {
        +String name
        +int age
        +makeSound()
    }
    class Dog {
        +String breed
        +bark()
    }
    Animal <|-- Dog
```

### Rules

- Start with `#| mermaid` comment
- Add chunk options with `#|` prefix (e.g., `#| eval: true`)
- Diagram content follows **without `#` prefix**
- Chunk ends at first blank line or comment line
- Options are automatically converted to Quarto format (`%%|`)

### Complete Example

See the complete example file with multiple diagram types:

```r
# Locate the Mermaid example file
mermaid_file <- system.file("examples", "example_mermaid.R", package = "quartify")

# Convert and render
rtoqmd(mermaid_file, render = TRUE, open_html = TRUE)
```

::: {.alert .alert-info}
**Tip:** Use `install_quartify_snippets()` to add a `mermaid` snippet. Type `mermaid` + `Tab` to insert the diagram template.

**More Mermaid Resources:**

- [Mermaid Documentation](https://mermaid.js.org/)
- [Mermaid Live Editor](https://mermaid.live/) - Test your diagrams
- [Quarto Mermaid Guide](https://quarto.org/docs/authoring/diagrams.html)
:::

## Tabsets

Tabsets allow you to organize related content in interactive tabs, perfect for displaying alternative views, different analyses, or grouped visualizations. This feature uses the same intuitive syntax as callouts.

### Basic Syntax

To create a tabset:

1. **Start tabset container**: Use `# tabset` on its own line
2. **Define each tab**: Use `# tab - Tab Title` on its own line
3. **Add content**: Add comments and code for each tab
4. **Automatic closing**: Tabset closes at the next section header

**Syntax in R script:**

```r
# tabset
# tab - Summary Statistics
# Here are the basic summary statistics for the iris dataset:
summary(iris)
```

**Resulting Quarto:**

```markdown
::: {.panel-tabset}

## Summary Statistics

Here are the basic summary statistics for the iris dataset:

```{r}
summary(iris)
```

:::
```

### Rules

- `# tabset` must be on its own line (with optional leading/trailing whitespace)
- Each tab starts with `# tab - Title` where `Title` is the tab label
- Tab content includes all comments and code until the next tab or tabset end
- Tabsets automatically close at RStudio section headers (`##`, `###`, `####`)
- You can have multiple tabsets in the same document
- Empty lines within tabs are preserved

### Complete Example

See the complete example file with multiple tabsets:

```r
# Locate the tabset example file
tabset_file <- system.file("examples", "example_tabset.R", package = "quartify")

# Convert and render
rtoqmd(tabset_file, render = TRUE, open_html = TRUE)
```

::: {.alert .alert-info}
**Tip:** Use `install_quartify_snippets()` to add a `tabset` snippet. Type `tabset` + `Tab` to insert the tabset template.

**More Tabset Resources:**

- [Quarto Tabset Documentation](https://quarto.org/docs/interactive/layout.html#tabset-panel)
- Combine tabsets with callouts for rich, organized documentation
:::

## Comment Rules Summary

| Type | Syntax | Result | Example |
|------|--------|--------|---------|
| **Level 2 Header** | `## Title ####` | Markdown `## Title` | `## Data Analysis ####` |
| **Level 3 Header** | `### Title ====` | Markdown `### Title` | `### Preprocessing ====` |
| **Level 4 Header** | `#### Title ----` | Markdown `#### Title` | `#### Remove NA ----` |
| **Standalone Comment** | `# Text` | Plain text paragraph | `# This filters the data` |
| **Callout** | `# callout-TYPE - Title` | Quarto callout block | `# callout-note - Important` |
| **Mermaid Diagram** | `#| mermaid` | Mermaid diagram chunk | `#| mermaid` + content |
| **Tabset** | `# tabset`, `# tab - Title` | Tabbed panel | `# tabset` + tabs |
| **Code** | No `#` prefix | R code chunk | `iris %>% filter(...)` |
| **Inline Comment** | `  # Text` in code | Stays in code chunk | `iris %>% # comment` |

::: {.alert .alert-danger}
<span style="color: red;">**Critical rules to avoid errors:**

1. **Spacing**: Always include a space after `#` for comments and section headers
2. **Trailing symbols**: Section headers MUST have at least 4 trailing symbols
3. **Symbol flexibility**: You can use `#`, `=`, or `-` interchangeably for trailing symbols, but following the RStudio convention is recommended
4. **Roxygen comments**: Lines starting with `#'` are ignored (for package development)
5. **Comment context**: Comments at the start of a line become text; comments within code stay in code blocks
6. **Keyboard shortcut**: Use `Ctrl+Shift+C` (Windows/Linux) or `Cmd+Shift+C` (Mac) to comment/uncomment lines quickly</span>
:::

## Code Quality Integration

quartify integrates with `styler` and `lintr` to provide automated code quality checks and formatting in your documentation.

### Using styler

The `use_styler` parameter automatically formats your code according to the tidyverse style guide:

```{r eval=FALSE}
# Create a script with style issues
temp_script <- tempfile(fileext = ".R")
writeLines(c(
  "x = 3  # Should use <- instead of =",
  "y<-2   # Missing spaces",
  "z <- mean(c(1,2,3))  # Missing spaces in function call"
), temp_script)

# Convert with styler
rtoqmd(temp_script, "styled_output.qmd", use_styler = TRUE)
```

The generated document will include tabsets showing:
- **Original Code**: Your code as written
- **Styled Code**: The tidyverse-formatted version

### Using lintr

The `use_lintr` parameter runs code quality checks:

```{r eval=FALSE}
# Convert with both styler and lintr
rtoqmd(temp_script, "quality_output.qmd", 
       use_styler = TRUE, 
       use_lintr = TRUE)
```

The tabsets will include:
- **Original Code**: Your code
- **Styled Code**: Formatted version
- **Lint Issues**: List of style violations and suggestions

### Applying styler to source files

Use `apply_styler = TRUE` to modify your source R script directly:

```{r eval=FALSE}
# This will modify the original R file
rtoqmd("my_script.R", "output.qmd", 
       use_styler = TRUE, 
       apply_styler = TRUE)
```

::: {.alert .alert-warning}
**Warning:** `apply_styler = TRUE` modifies your original R script file. Consider using version control or backups before using this option.
:::

### Example Output

Try the included example:

```{r eval=FALSE}
example_file <- system.file("examples", "example_code_quality.R", 
                           package = "quartify")
rtoqmd(example_file, "quality_demo.qmd", 
       use_styler = TRUE, 
       use_lintr = TRUE,
       render_html = TRUE,
       open_html = TRUE)
```

## Tips and Best Practices

1. **Start with structure**: Define your section headers first to create the document outline
2. **Use consistent levels**: Follow a logical hierarchy (2 → 3 → 4, don't skip levels)
3. **Add explanatory text**: Use regular comments to explain what your code does
4. **Group related code**: Keep related operations together; they'll be grouped into the same code block
5. **Test incrementally**: Start with a small script to see how the conversion works
6. **Navigate easily**: In RStudio, use the document outline (Ctrl+Shift+O) to see your structure
7. **Comment liberally**: More comments = better documentation in the final Quarto document
8. **Use code quality checks**: Enable `use_styler` and `use_lintr` for teaching materials or code reviews

## Conclusion

`quartify` makes it easy to transform your R scripts into professional Quarto documents without manual reformatting. By following RStudio code section conventions, you can automatically generate well-structured, reproducible documentation from your existing code with proper navigation hierarchy.
