---
title: "Roxygen Documentation Example"
author: "Quartify Team"
date: "2025-12-09"
format:
  html:
    embed-resources: true
    code-fold: false
    number-sections: true
toc: true
toc-title: Sommaire
toc-depth: 4
toc-location: left
execute: 
  eval: false
  echo: true
output:
  html_document:
  output-file: example_roxygen.html
---

## Mathematical Functions

::: {.callout-note}

## Documentation - add_numbers

Add Two Numbers

This function takes two numeric values and returns their sum.

@param x A numeric value.
@param y A numeric value.

@return A numeric value representing the sum of `x` and `y`.

@examples
add_numbers(3, 5)   # returns 8
add_numbers(-2, 7)  # returns 5

@export

:::

```{r}
# Lignes 21-23
add_numbers <- function(x, y) {
  x + y
}
```

::: {.callout-note}

## Documentation - multiply_numbers

Multiply Two Numbers

This function multiplies two numeric values.

@param a First number
@param b Second number
@return The product of a and b
@examples
multiply_numbers(4, 5)  # returns 20
@export

:::

```{r}
# Lignes 35-37
multiply_numbers <- function(a, b) {
  a * b
}
```

## String Functions

::: {.callout-note}

## Documentation - to_upper

Convert Text to Uppercase

Takes a character string and converts all letters to uppercase.

@param text A character string to convert
@return The input string in uppercase
@examples
to_upper("hello world")  # returns "HELLO WORLD"
@export

:::

```{r}
# Lignes 50-52
to_upper <- function(text) {
  toupper(text)
}
```

Regular comment without roxygen

```{r}
# Lignes 55-56
result <- add_numbers(10, 20)
print(result)
```

